---
title: "Homework 1"
author: "YOUR NAME HERE"
format:
  html:
    self-contained: true
editor_options: 
  chunk_output_type: console
---

## Instructions

- Answer every question directly in this `.qmd` file.
- Written answers can go in plain text below each question — show your reasoning, not just a final number.
- Questions with a code chunk require you to actually run code, not just report a number you found by hand.
- When finished, **render this file to HTML** and submit the `.html` file.

```{r setup, message=FALSE}
library(tidyverse)
library(openintro)
data(nycflights)
```

------------------------------------------------------------------------

# Part 1: Probability (OpenIntro Statistics, Ch. 3)

## Q1. Probability distributions

A six-sided die is rolled once.

a.  Write out the probability distribution for the outcome (i.e., a table listing each possible outcome and its probability).
b.  Confirm this distribution satisfies the three rules for a valid probability distribution.
c.  What is the probability of rolling an even number? Show your work.
d.  **Code:** Use R as a calculator to confirm your answer to (c) — e.g. by adding up the relevant probabilities from your table.

```{r q1}
# TODO: confirm your probability calculation

```

::: {.callout-note title="A quick note on independence"}
Two events $A$ and $B$ are **independent** if knowing that one occurred tells you nothing about the probability of the other — that is, if

$$P(A \mid B) = P(A)$$

Concretely: compare $P(A)$ (the probability of $A$ overall, ignoring $B$) to $P(A \mid B)$ (the probability of $A$ *given* that $B$ happened). If they're the same (or very close), the events are independent. If $P(A \mid B)$ is noticeably different from $P(A)$, then knowing $B$ happened *does* change how likely $A$ is — so the events are **dependent**, not independent.
:::


## Q2. Conditional probability from a table

A sample of 150 students was asked whether they own a car and whether they live on campus:

|                    | Owns a car | No car | Total |
|--------------------|:---:|:---:|:---:|
| **Lives on campus** | 18 | 72 | 90 |
| **Lives off campus** | 42 | 18 | 60 |
| **Total**          | 60 | 90 | 150 |

a.  What is the probability that a randomly selected student owns a car?
b.  What is the probability that a student owns a car, *given* that they live off campus?
c.  Based on (a) and (b), does car ownership appear to be independent of living situation? Explain using the definition of independence.
d.  **Code:** Enter this table in R (as a matrix or data frame) and use `prop.table()` to confirm your answer to part (b).

```{r q2}
# TODO: enter the table and confirm the conditional probability

```

## Q3. Disjoint vs. independent

Consider drawing one card from a standard 52-card deck.

a.  Are the events "drawing a face card" (jack, queen, king) and "drawing a heart" disjoint? Explain why or why not.
b.  Calculate $P(\text{face card OR heart})$ using the general addition rule.
c.  **Code:** Confirm your answer to (b) using R as a calculator.

```{r q3}
# TODO: confirm your addition-rule calculation

```

------------------------------------------------------------------------

# Part 2: Exploring Categorical Data (IMS, Ch. 4)

We'll use the `nycflights` dataset from Friday's lab for the rest of this assignment.

## Q4. Summarizing a categorical variable

a.  Use `count()` (or `group_by()` + `summarise()`) to find how many flights each `carrier` had in this dataset.
b.  Add a column showing each carrier's proportion of total flights.
c.  Which carrier had the most flights? What proportion of all flights does that represent?

```{r q4}
# TODO: count flights by carrier and compute proportions

```

## Q5. Visualizing a categorical variable

a.  Make a bar plot of `origin` (which airport flights departed from) using `ggplot2`.
b.  In a sentence or two, describe what the plot shows — is one airport clearly busier than the others in this sample?

```{r q5}
# TODO: bar plot of origin

```

## Q6. Two categorical variables together

Recall from Monday's lecture that we can compare row proportions in a contingency table to look for a relationship between two categorical variables.

a.  Create a new variable `dep_type` that labels each flight `"on time"` if `dep_delay < 5` and `"delayed"` otherwise (hint: `mutate()` + `ifelse()`).
b.  Make a **filled** (standardized/proportional) bar plot of `dep_type` by `origin` (hint: `geom_bar(position = "fill")`).
c.  Based on your plot, does on-time departure rate appear to depend on which airport a flight departs from? Which airport looks best/worst?

```{r q6}
# TODO: create dep_type and make the filled bar plot

```

------------------------------------------------------------------------

# Part 3: Exploring Numerical Data (IMS, Ch. 5)

## Q7. Summary statistics

a.  Use `summarise()` to compute the mean, median, standard deviation, and IQR of `dep_delay`.
b.  Given what you found, would you expect this distribution to be symmetric or skewed? Explain your reasoning using the relationship between the mean and median from Wednesday's lecture.

```{r q7}
# TODO: compute summary statistics for dep_delay

```

## Q8. Visualizing shape

a.  Make a histogram of `dep_delay`.
b.  Describe the shape of the distribution: is it unimodal or multimodal? Symmetric or skewed (and if skewed, which direction)? Are there any potential outliers?
c.  Does the shape match what you predicted in Q7(b)?

```{r q8}
# TODO: histogram of dep_delay

```

## Q9. Comparing a numerical variable across groups

a.  Make side-by-side boxplots of `dep_delay` by `origin`.
b.  Which origin airport has the highest *median* departure delay? Which has the most *variability* (widest spread/IQR)?
c.  Based on what you learned about robust statistics, would you trust the mean or the median more to describe "typical" delay at each airport? Explain.

```{r q9}
# TODO: side-by-side boxplots of dep_delay by origin

```
