---
title: "Homework: Hypothesis Testing, Confidence Intervals \\& Proportions"
subtitle: "IMS Chapters 11-14, 16-17"
author: "YOUR NAME HERE"
format:
  html:
    self-contained: true
    toc: true
---

## Instructions

- Show your work for every computation, not just a final number.
- Questions with a code chunk require you to actually run code.
- Render this file to HTML and submit the `.html` file.

```{r setup, message=FALSE}
library(tidyverse)
library(openintro)
data(avandia)
data(heart_transplant)
```

------------------------------------------------------------------------

# Chapter 11: Hypothesis Testing with Randomization

## Q1. Identify the parameter, II

For each situation below, state whether the parameter of interest is a **mean** or a **proportion**.

a.  A poll shows that 64% of Americans personally worry a great deal about federal spending and the budget deficit.
b.  A survey reports that local TV news revenue increased 17% over a two-year period, while newspaper revenue decreased 6.4% over the same period.
c.  High school and college students are asked whether they use geolocation services on their smartphones.
d.  Smartphone users are asked whether they use a web-based taxi service.
e.  Smartphone users are asked how many times they used a web-based taxi service over the last year.

## Q2. Identify hypotheses, II

Write the null and alternative hypotheses in words **and** in symbols for each situation.

a.  Since 2008, California chain restaurants have been required to display calorie counts on menus. Before calorie counts were displayed, average calorie intake at a restaurant was 1100 calories. A nutritionist wants to know if average calorie intake has changed since calorie counts started being displayed.
b.  The average GRE Verbal Reasoning score was 462 for test-takers between July 2004 and June 2007. In 2021, the average was slightly higher. Does this provide convincing evidence that the average GRE Verbal score has changed since 2021?

## Q3. Side effects of Avandia

Rosiglitazone (Avandia) and Pioglitazone (Actos) are both type-2 diabetes medications. In a retrospective study of 227,571 Medicare beneficiaries, 2,593 of 67,593 Rosiglitazone users and 5,386 of 159,978 Pioglitazone users had serious cardiovascular problems. The `avandia` dataset (in the `openintro` package) contains this data.

a.  What proportion of all patients in the study had cardiovascular problems?
b.  If treatment and cardiovascular outcome were independent, how many patients in the Rosiglitazone group would we *expect* to have had cardiovascular problems? Compare this to the number actually observed.
c.  **True or false, with explanation:** "Since the rate of incidence was higher for the Rosiglitazone group (3.8% vs. 3.4%), Rosiglitazone *causes* cardiovascular problems." Be careful — a statement's conclusion can be correct while the reasoning behind it is still wrong.
d.  **Code:** Using the `avandia` dataset, run a randomization test (shuffle the treatment labels, recompute the difference in proportions, repeat many times) to evaluate whether treatment and cardiovascular outcome are independent. Report your simulated p-value and conclusion.

```{r q3}
# TODO: randomization test using the avandia dataset

```

------------------------------------------------------------------------

# Chapter 12: Bootstrapping

## Q4. Cyberbullying rates

A survey found that 54% to 64% of teens reported experiencing cyberbullying (a 95% confidence interval).

a.  A newspaper claims "a majority of teens have experienced cyberbullying." Is this claim supported by the interval? Explain.
b.  A researcher conjectured that 70% of teens have experienced cyberbullying. Is this claim supported by the interval? Explain.
c.  Without recalculating, would the researcher's 70% claim from part (b) be supported by a **90%** confidence interval instead? Explain your reasoning (think about how interval width changes with confidence level).

## Q5. Chronic illness, bootstrapped

The 2012 Pew Research Foundation report found that 45% of a sample of 3,014 US adults live with one or more chronic conditions.

a.  **Code:** Simulate a dataset consistent with this study (3,014 observations, 45% "yes") and construct a 92% bootstrap percentile confidence interval for the true proportion (resample with replacement, 1000+ times).
b.  Interpret your interval in the context of the study.
c.  Is 50% inside your interval? What does that tell you about whether a majority of US adults live with a chronic condition?

```{r q5}
# TODO: simulate the sample and bootstrap a 92% CI

```

------------------------------------------------------------------------

# Chapter 13: Mathematical Models

## Q6. Interpreting a Z-score from a sample proportion

Suppose you conduct a hypothesis test about a population proportion and calculate the Z-score to be 0.47.

For each statement below, say whether it is the **best interpretation** of this value. For every statement that is *not* a good interpretation, identify the statistical idea it's actually (incorrectly) describing.

a.  "The probability is 0.47 that the null hypothesis is true."
b.  "If the null hypothesis were true, the probability would be 0.47 of obtaining a sample proportion as far from the hypothesized value as observed."
c.  "The sample proportion is 0.47 standard errors greater than the hypothesized value of the population proportion."
d.  "The sample proportion is equal to 0.47 times the standard error."
e.  "The sample proportion is 0.47."

## Q7. Social media users and news, mathematical model

A 2022 poll found that 50% of US adults get news from social media sometimes or often, with a reported standard error of 0.5%.

a.  Construct a 99% confidence interval for the true proportion, and interpret it in context.
b.  **True or false, with explanation, for each:**
    i.  "The data provide statistically discernible evidence that more than half of US adults get news through social media sometimes or often" (using $\alpha=0.05$).
    ii. "Since the standard error is 0.5%, we can conclude that 99.5% of all US adults were included in the study."
    iii. "If we want to reduce the standard error of the estimate, we should collect less data."
    iv. "A 90% confidence interval for this same estimate will be *wider* than the 99% confidence interval."

## Q8. Quality control

An engineer at a factory samples 212 computer chips during a week of production and finds 27 defective.

a.  What is the point estimate for the true proportion of defective chips this week?
b.  Compute the standard error of this estimate using the mathematical model.
c.  The historical defect rate is 10%. Should the engineer be surprised by this week's observed rate? Support your answer with a hypothesis test.
d.  If the true population defect rate really is 10%, recompute the standard error using $p=0.10$ in place of $\hat p$. How much does the standard error change compared to part (b)?
e.  **Code:** Confirm your hypothesis test from part (c) using `prop.test()`.

```{r q8}
# TODO: prop.test() for the chip defect rate

```

------------------------------------------------------------------------

# Chapter 14: Decision Errors

## Q9. Testing for food safety

A food safety inspector investigates a restaurant after customer complaints about poor sanitation. If he decides the restaurant is in gross violation, its license to serve food will be revoked.

a.  Write the hypotheses (in words) for this scenario.
b.  What would a Type I error mean here?
c.  What would a Type II error mean here?
d.  Which error is more problematic for the *restaurant owner*? Explain.
e.  Which error is more problematic for the *diners*? Explain.
f.  As a diner, would you want the inspector to require "strong evidence" or "very strong evidence" of a violation before revoking a license? Explain your reasoning in terms of the two error rates.

## Q10. Which is higher?

For each pair of scenarios, state whether the value of interest is larger under scenario (i), scenario (ii), or equal in both. Briefly justify each answer.

a.  The standard error of $\hat p$ when (i) $n=100$ or (ii) $n=400$.
b.  The margin of error of a confidence interval when the confidence level is (i) 90% or (ii) 80%.
c.  The p-value for a Z-statistic of 2.5, calculated from (i) a sample of $n=50$ or (ii) a sample of $n=500$.
d.  The probability of a Type II error, when $H_A$ is true, at a discernibility level of (i) 0.05 or (ii) 0.10.

## Q11. True or false

Determine whether each statement is true or false. If false, explain how to correct it.

a.  If a given value is within a 95% confidence interval, it will also be within a 99% confidence interval built from the same data.
b.  Decreasing the discernibility level ($\alpha$) will increase the probability of making a Type I error.
c.  Suppose $H_0: p=0.5$ and we fail to reject $H_0$. Under this scenario, the true population proportion is 0.5.
d.  With large sample sizes, even small differences between the null value and the observed point estimate (the "effect size") will tend to be identified as statistically discernible.

## Q12. Online communication — spot the error, then run it correctly

A study suggests 60% of college students spend 10+ hours per week communicating online. You suspect this is wrong and sample 160 students from your dorm, finding 70% meet this description. A friend proposes:

$$H_0: \hat p = 0.60 \qquad H_A: \hat p \ne 0.70$$

a.  Identify every error in your friend's hypothesis statement. (Hint: think carefully about what symbol belongs in a hypothesis statement, and what value(s) belong on each side.)
b.  Write the correct hypotheses.
c.  **Code:** Actually run this hypothesis test (using either a randomization/simulation approach or `prop.test()`), and state your conclusion in context.

```{r q12}
# TODO: correctly test H0: p = 0.60 vs Ha: p != 0.60

```
