---
title: "Lab: Hypothesis Testing, Bootstrapping, Mathematical Models & Decision Errors"
format:
  html:
    self-contained: true
    toc: true
editor: visual
editor_options: 
  chunk_output_type: console
---

## Getting started

```{r setup, message=FALSE}
library(tidyverse)
library(openintro)
data(avandia)
data(cpr)
```

Today has two goals: (1) a quick refresher on the three methods you already know (randomization, bootstrapping, mathematical models), and (2) using **simulation** to actually *see* the ideas from Chapter 14 — Type I error rate, power, and why switching to a one-sided test after peeking at your data is a bad idea — rather than just reading about them.

------------------------------------------------------------------------

## Part 1: Quick review — randomization and `prop.test()`

Using the `avandia` dataset (diabetes drug side-effect study from homework):

```{r}
glimpse(avandia)
table(avandia)
```

**Try it:**

1.  State $H_0$ and $H_A$ for a two-sided test of whether `rosi` (treatment) and cardiovascular outcome are independent.
2.  Run a randomization test (shuffle, recompute, repeat) **and** a `prop.test()`, and compare the two p-values.

```{r part1}
# TODO: randomization test
n_control <- sum(avandia$treatment == "Pioglitazone")
n_trt <- sum(avandia$treatment == "Rosiglitazone")

trt_prop <- filter(avandia, treatment == "Rosiglitazone" & cardiovascular_problems == "yes") %>% nrow() / nrow(avandia)

control_pop <- filter(avandia, treatment == "Pioglitazone" & cardiovascular_problems == "yes") %>% nrow() / nrow(avandia)

obs_diff <- trt_prop - control_pop

sims <- map_dbl(1:10000, ~ {
 shuffle <- sample(avandia$cardiovascular_problems)
 phat_trt <- mean(shuffle[1:n_trt] == "yes")
 phat_control <- mean(shuffle[(n_trt+1):(n_trt+n_control)] == "yes")
 return(phat_trt - phat_control)
}
)

mean(abs(sims) >= abs(obs_diff))

# TODO: prop.test()
prop.test(n = c(n_trt, n_control), x = c(2593, 5386))
```

## Part 2: Quick review — bootstrapping

Using the same `avandia` data, treat the Rosiglitazone group alone as your sample of interest.

**Try it:** Construct a 95% bootstrap percentile confidence interval for the true proportion of Rosiglitazone users who experience cardiovascular problems.

```{r part2}
# TODO: bootstrap CI for the Rosiglitazone group's event rate

```

------------------------------------------------------------------------

## Part 3: Seeing $\alpha$ as the Type I error rate

Here's the key idea we'll test empirically: if $H_0$ is *actually true*, and you run a hypothesis test at $\alpha = 0.05$, you should incorrectly reject $H_0$ about 5% of the time — no more, no less. Let's check.

```{r}
set.seed(2026)
n_sim <- 1000
n <- 40
p_null <- 0.5   # we will simulate data where this null is EXACTLY true

successes <- rbinom(n_sim, size = n, prob = p_null)
pvals <- sapply(successes, function(x) prop.test(x, n, p = p_null)$p.value)

mean(pvals < 0.05)
```

**Discuss:** Is this close to 0.05? If it's noticeably *lower*, recall what we learned about `prop.test()`'s default continuity correction — try re-running with `correct = FALSE` inside the `prop.test()` call above and see if the empirical rate moves closer to 0.05.

```{r part3-nocorrect}
# TODO: repeat using correct = FALSE and compare
successes <- rbinom(n_sim, size = n, prob = p_null)
pvals <- sapply(successes, function(x) prop.test(x, n, p = p_null, correct = FALSE)$p.value)

mean(pvals < 0.05)
```

**Try it:** Repeat the whole simulation using $\alpha = 0.01$ instead of $0.05$. What empirical Type I error rate do you get? Does it match what you'd expect?

## Part 4: Seeing power (and 1 − Type II error rate)

Now simulate a situation where $H_A$ is **actually true** — the real proportion is 0.65, but we're still testing against a null of 0.5. How often does our test correctly catch that?

```{r}
set.seed(2026)
n_sim <- 1000
n <- 40
p_true <- 0.65
p_null <- 0.5

successes <- rbinom(n_sim, size = n, prob = p_true)
pvals <- sapply(successes, function(x) prop.test(x, n, p = p_null)$p.value)

power_est <- mean(pvals < 0.05)
power_est
1-power_est
```

```{r}
tibble(props_true = successes/n,
       props_null = rbinom(n_sim, size = n, prob = p_null)/n) %>% 
  pivot_longer(cols = everything(), names_to = "null_truth", values_to = "props") %>% 
  ggplot(aes(props, fill = null_truth)) +
  geom_histogram() 
```

This proportion — how often we correctly reject a false $H_0$ — is an estimate of **power**. The complement, `1 - power_est`, estimates the **Type II error rate**.

**Try it:** Recompute power using $n=100$ instead of $n=40$ (keep $p_{true}=0.65$, $p_{null}=0.5$). Did power go up or down? Does this match what lecture said about sample size and power?

```{r part4}
# TODO: recompute power with n = 100
set.seed(2026)
n_sim <- 1000
n <- 100
p_true <- 0.65
p_null <- 0.5

successes <- rbinom(n_sim, size = n, prob = p_true)
pvals <- sapply(successes, function(x) prop.test(x, n, p = p_null)$p.value)

power_est <- mean(pvals < 0.05)
power_est
1-power_est
```

```{r}
tibble(props_true = successes/n,
       props_null = rbinom(n_sim, size = n, prob = p_null)/n) %>% 
  pivot_longer(cols = everything(), names_to = "null_truth", values_to = "props") %>% 
  ggplot(aes(props, fill = null_truth)) +
  geom_histogram() 
```

**Try it:** Now go back to $n=40$, and try `p_true = 0.55` instead of `0.65` (a smaller true effect). What happens to power? Connect this to what lecture said about *effect size* and power.

```{r part4b}
# TODO: recompute power with a smaller true effect
set.seed(2026)
n_sim <- 1000
n <- 40
p_true <- 0.55
p_null <- 0.5

successes <- rbinom(n_sim, size = n, prob = p_true)
pvals <- sapply(successes, function(x) prop.test(x, n, p = p_null)$p.value)

power_est <- mean(pvals < 0.05)
power_est
1-power_est
```

```{r}
tibble(props_true = successes/n,
       props_null = rbinom(n_sim, size = n, prob = p_null)/n) %>% 
  pivot_longer(cols = everything(), names_to = "null_truth", values_to = "props") %>% 
  ggplot(aes(props, fill = null_truth)) +
  geom_histogram() 
```

```{r part4b}
# TODO: recompute power with a smaller true effect
set.seed(2026)
n_sim <- 1000
n <- 100
p_true <- 0.55
p_null <- 0.5

successes <- rbinom(n_sim, size = n, prob = p_true)
pvals <- sapply(successes, function(x) prop.test(x, n, p = p_null)$p.value)

power_est <- mean(pvals < 0.05)
power_est
1-power_est
```

```{r}
tibble(props_true = successes/n,
       props_null = rbinom(n_sim, size = n, prob = p_null)/n) %>% 
  pivot_longer(cols = everything(), names_to = "null_truth", values_to = "props") %>% 
  ggplot(aes(props, fill = null_truth)) +
  geom_histogram() 
```

## Part 5: Why you can't switch sides after peeking

Lecture argued that if you always pick whichever *one-sided* test matches the direction your data happens to lean, your true Type I error rate silently doubles (roughly $5\% + 5\% = 10\%$, not 5%). Let's simulate this directly.

```{r}
set.seed(2026)
n_sim <- 5000
n <- 40
p_null <- 0.5   # null is TRUE in this simulation -- there is no real effect

successes <- rbinom(n_sim, size = n, prob = p_null)
phat <- successes / n

# The PROPER approach: pre-register a two-sided test, always
pvals_two_sided <- sapply(successes, function(x) prop.test(x, n, p = p_null)$p.value)
type1_two_sided <- mean(pvals_two_sided < 0.05)

# The "cheating" approach: after seeing the data, pick whichever one-sided
# alternative matches the direction phat happens to lean
pvals_switched <- sapply(seq_along(successes), function(i) {
  alt <- if (phat[i] > p_null) "greater" else "less"
  prop.test(successes[i], n, p = p_null, alternative = alt)$p.value
})
type1_switched <- mean(pvals_switched < 0.05)

type1_two_sided
type1_switched
```

**Discuss:** How do these two empirical error rates compare? Does the "switching" strategy's error rate land close to lecture's predicted \~10%? Why is this a problem in practice — what would it look like if a researcher did this without realizing it?

------------------------------------------------------------------------

## Part 6: In-class exercise

Using the `cpr` dataset (the blood-thinner study):

1.  State $H_0$/$H_A$ for a two-sided test comparing survival rates between `treatment` and `control`.
2.  Run the test using `prop.test()`.
3.  Suppose you wanted to know: if the *true* difference in survival rates were as large as what you observed in this sample, how much power would a study of this exact size ($n_{treatment}=40$, $n_{control}=50$) have had to detect it? Design and run a simulation (adapting the code from Part 4) to estimate this.
4.  In 2-3 sentences, explain what your power estimate tells you about how convincing a "fail to reject" result would have been, *if* that's what this study had found.

```{r part6}
# TODO: hypothesis test

# TODO: power simulation for this specific study design

```
