---
title: "Exploratory Analysis"
subtitle: "Putting categorical and numerical summaries into code"
format:
  html:
    self-contained: true
    toc: true
editor: visual
editor_options: 
  chunk_output_type: console
---

## Getting started

This lab picks up where our RStudio/tidyverse basics left off. This week we saw how to *describe* categorical data (bar plots, contingency tables) and numerical data (histograms, boxplots, mean/median, variance/SD) by hand and by sight. Today we'll build those same summaries and plots in code, using one dataset throughout: `nycflights`, a random sample of domestic flights departing the three major NYC airports in 2013.

```{r setup, message=FALSE}
library(tidyverse)
library(openintro)
data(nycflights)
```

Take a first look:

```{r}
glimpse(nycflights)
```

Some variables you'll use today:

- **Categorical:** `carrier` (airline), `origin` (JFK/LGA/EWR), `dest`, `month`
- **Numerical:** `dep_delay`, `arr_delay`, `distance`, `air_time`

------------------------------------------------------------------------

## Part 1: Wrangling review

Before we visualize, a quick refresher on the verbs that get data into plotting shape.

### Filtering

Keep only the rows you care about with `filter()`. Flights headed to Los Angeles:

```{r}
lax_flights <- nycflights |>
  filter(dest == "LAX")
```

Multiple conditions — flights to San Francisco in February:

```{r}
sfo_feb_flights <- nycflights |>
  filter(dest == "SFO", month == 2)
### also or |
```

### Creating new variables

`mutate()` adds a column. Classify each flight as on-time or delayed:

```{r}
nycflights <- nycflights |>
  mutate(dep_type = ifelse(dep_delay < 5, "on time", "delayed"))
```

### Grouping and summarizing

`group_by()` + `summarise()` is how you get one row of stats per category:

```{r}
nycflights |>
  group_by(origin) |>
  summarise(
    median_dd = median(dep_delay),
    iqr_dd    = IQR(dep_delay),
    n_flights = n()
  )
```

### Sorting

`arrange(desc(...))` puts the largest values first:

```{r}
nycflights |>
  group_by(month) |>
  summarise(mean_dd = mean(dep_delay)) |>
  arrange(desc(mean_dd))
```

**Try it:** Which origin airport has the best on-time departure *rate* (proportion of flights with `dep_type == "on time"`)? Write the pipe chain to find out.

```{r}
nycflights %>% 
  mutate(ontime_num = as.numeric(dep_type == "on time")) %>% 
  group_by(origin) %>% 
  summarise(ontime_rate = sum(ontime_num)/n(),
            ontime_rate_mean = mean(ontime_num)) %>% 
  arrange(desc(ontime_rate))

nycflights %>% 
  group_by(origin) %>% 
  summarise(ontime_rate = mean(dep_type == "on time")) %>% 
  arrange(desc(ontime_rate))
```


------------------------------------------------------------------------

## Part 2: Visualizing — matching the plot to the variable(s)

Every `ggplot2` call follows the same recipe: `ggplot(data, aes(...)) + geom_...()`. The **geom** you choose depends on how many variables you have and whether they're categorical or numerical — this is the same decision tree from Monday and Wednesday's lectures, just now in code.

### One categorical variable → bar plot

```{r}
ggplot(data = nycflights, aes(x = carrier)) +
  geom_bar()
```

Relative frequency instead of counts — divide by the total, or let `ggplot` do proportions directly:

```{r}
ggplot(data = nycflights, aes(x = carrier, y = after_stat(count) / sum(after_stat(count)))) +
  geom_bar() +
  ylab("proportion")
```

### One numerical variable → histogram, density, or boxplot

Histogram — the shape of a single distribution:

```{r}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram()
```

Binwidth changes the story (recall Wednesday's discussion of bin width):

```{r}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram(binwidth = 15)
```

Density plot — a smoothed version of the histogram:

```{r}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_density()
```

Boxplot — center, spread, and outliers at a glance:

```{r}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_boxplot()
```

### Two categorical variables → stacked or grouped bar plot

Is on-time performance related to which airport you fly out of?

```{r}
ggplot(data = nycflights, aes(x = origin, fill = dep_type)) +
  geom_bar()
```

As proportions instead of counts (a **standardized/filled** bar plot — same idea as the row proportions we calculated by hand Monday):

```{r}
ggplot(data = nycflights, aes(x = origin, fill = dep_type)) +
  geom_bar(position = "fill") +
  ylab("proportion")
```

Side-by-side instead of stacked:

```{r}
ggplot(data = nycflights, aes(x = origin, fill = dep_type)) +
  geom_bar(position = "dodge")
```

### One numerical + one categorical → side-by-side boxplots

Do departure delays vary by origin airport?

```{r}
ggplot(data = nycflights, aes(x = origin, y = dep_delay)) +
  geom_boxplot()
```

### Two numerical variables → scatterplot

Does flight distance predict air time?

```{r}
ggplot(data = nycflights, aes(x = distance, y = air_time)) +
  geom_point()
```

Add a trend line to see the relationship more clearly:

```{r}
ggplot(data = nycflights, aes(x = distance, y = air_time)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE)
```

### Bonus: splitting a plot by a third variable (faceting)

If you want to see the distance/air-time relationship separately for each origin airport, instead of coloring points you can facet into small multiples:

```{r}
ggplot(data = nycflights, aes(x = distance, y = air_time)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ origin)

ggplot(data = nycflights, aes(x = distance, y = air_time, color = origin)) +
  geom_point() +
  geom_smooth() 
```

------------------------------------------------------------------------

## Part 3: In-class exercise

Using `nycflights`, answer the following. For each, decide what type of variable(s) you have first — that tells you what plot to reach for.

1.  Make a plot showing the distribution of `arr_delay` for flights **only** headed to `"ORD"`. Describe the shape (modality, skew, outliers).

```{r}
nycflights %>% 
  filter(dest == "ORD") %>% 
  ggplot(aes(x = arr_delay)) +
  geom_histogram() 

nycflights %>% 
  filter(dest == "ORD") %>% 
  ggplot(aes(x = arr_delay)) +
  geom_density()
```

2.  Compare `arr_delay` across the three origin airports with a single plot. Which airport looks most consistent (least variable)?
3.  Is there a relationship between `dep_delay` and `arr_delay`? Make an appropriate plot and describe what you see.
4.  Make a bar plot showing how many flights each `carrier` had that were `"delayed"` vs. `"on time"`. Which carrier has the worst on-time rate?
5.  **Open-ended:** Come up with your own question about `nycflights` that involves at least one numerical and one categorical variable, then answer it with a plot.

*(If you'd rather work with a different dataset for #5, `cubs_all_time.csv` also works well here — swap in `read_csv("cubs_all_time.csv")` for the `data` argument in any of the geoms above.)*
